---
title: "Cell type identification"
output: github_document
---

Created by: Philip Lijnzaad

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=9, fig.height=6) 
```

# Overview

In any single-cell experiment where mixed populations are expected, a very important task is to identify the cell types (and/or cell states) that are present. In this practical, we will look at two tools for doing this: SingleR ([Aran et al. 2019](https://www.nature.com/articles/s41590-018-0276-y)) and CHETAH ([de Kanter et al., under review](http://dx.doi.org/10.1101/558908)).

# Datasets

We will try to classify the ovarian ascites data from [Schelker et al. 2017](https://www.nature.com/articles/s41467-017-02289-3) using, as a reference, the different reference data sets from SingleR and CHETAH. Let's first load the various pacckages.

```{r packages}
suppressMessages(require(SingleCellExperiment))
suppressMessages(require(Seurat))
suppressMessages(require(CHETAH))
suppressMessages(require(Matrix))
suppressMessages(require(SingleR))
suppressMessages(require(cowplot))
```

Let's get the data; if you have (or downloaded) the data into folder
`session-celltypeid_files` then load them with:


```{r load}
#load expression matrix

data.dir <- "~/git/excelerate-scRNAseq/session-celltypeid/session-celltypeid_files"
file <- paste(sep="/",data.dir,"ovarian1200.rds")
ovarian <- readRDS(file=paste(sep="/",data.dir,"ovarian.rds") )

#look at the cell type annotation
table(colData(deng)$cell_type2)

```



### Session info

```{r}
sessionInfo()
```





